---
title: "COVID project"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "2023-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Include the required libraries
```{r library}
library(tidyverse)
library(lubridate)
```

# Importing data

The present study entails the acquisition of COVID-19 data from a reliable and reputable source. 

```{r import}
# Record urls
urls <- c('https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv', 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv', 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv', 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv')
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

# Import 4 csvs
US_cases <- read.csv(urls[1])
US_deaths <- read.csv(urls[2])
global_cases <- read.csv(urls[3])
global_deaths <- read.csv(urls[4])

# Import population data
uid <- read.csv(uid_lookup_url) %>%
  select(c(Province_State, Country_Region, Population)) %>%
  mutate_all(~na_if(., ''))
```

# Tidying and transforming data

Tidy and transform imported data, create new columns for analysis.



```{r tidy}

# Create a data frame for global cases
global_cases <- global_cases %>%
  # Remove some columns
  select(-c(Lat, Long)) %>%
  # show date and cases in 2 columns
  pivot_longer(cols = -c(Province.State, 
                        Country.Region),
               names_to = "date",
               values_to = "cases")

# Create a data frame for global deaths
global_deaths <- global_deaths %>%
  select(-c(Lat, Long)) %>%
  pivot_longer(cols = -c(Province.State, 
                        Country.Region),
               names_to = "date",
               values_to = "deaths")

# Join global_cases and global_deaths to create final global data frame
global <- global_cases %>%
  full_join(global_deaths) %>%
  # Change date in correct format
  transform(date = gsub("X","",as.character(date))) %>%
  # Transform date from character to date object
  mutate(date = mdy(date)) %>%
  # Change all blank data to NA
  mutate_all(~na_if(., '')) %>%
  # Create a Combined_Key column
  unite("Combined_Key",
        c(Province.State, Country.Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE) %>%
  rename(Country_Region = Country.Region,
         Province_State = Province.State) %>%
  # Join population data to the data frame
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  # Select columns for analysis
  select(Province_State, Country_Region, date, 
         cases, deaths, Population, Combined_Key)

# Create a data frame for US cases
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  select(-c(Lat, Long_))

# Create a data frame for US deaths
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  select(-c(Lat, Long_))

# Join US_cases and US_deaths to create final US data frame
US <- US_cases %>%
  full_join(US_deaths) %>%
  transform(date = gsub("X","",as.character(date))) %>%
  mutate(date = mdy(date))

# Show summary for global data
summary(global)

# Show summary for US data
summary(US)
```
# Data Analysis

## Canada data analysis

Retrieve Canadian data by extracting it from the global data set.

```{r analysis_Canada1}
# Extract Canada data from global data
Canada <- global %>%
  filter(Country_Region == "Canada")

# Group Canada data by province
Canada_by_province <- Canada %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  ungroup()

# Calculate Canada's national data
Canada_totals <- Canada_by_province %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population,  na.rm = TRUE)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  ungroup()

# Plot a log-scale line graph that displays the cumulative COVID-19 cases and deaths in Canada over time
Canada_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in Canada", y = NULL)
```
The presented graph illustrates the trend of cumulative COVID-19 cases and related fatalities in Canada. The recorded cases begin from February 2020 and exhibit a sharp upward trajectory before stabilizing at a plateau.

## Canadian province analysis

The process of extracting provincial data from the national dataset is undertaken, followed by the development of a function for generating a graph of the number of COVID-19 cases over time for each province.
```{r analysis_province}
# Create a function to plot a graph for Province
plot_province <- function(province) {
  Canada_by_province %>%
  filter(Province_State == province) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", province), y = NULL)
}

# Plot a graph for Ontario
plot_province("Ontario")

# Plot a graph for British Columbia
plot_province("British Columbia")
```

The initial graph demonstrates the progression of COVID-19 cases in Ontario, while the subsequent graph displays the same for British Columbia. Although both graphs exhibit comparable trends to the national trend, minor variations exist within the first year of data.

## New cases analysis

A calculation of new COVID-19 cases was performed for the Canadian data set, with the results tabulated on a monthly basis.
```{r analysis_Canada2}

Canada_totals %>%
  # Extract year and month part from date
  mutate(year_month = format_ISO8601(date, precision = "ym")) %>%
  # Calculate new cases and deaths 
  mutate(new_cases = cases - lag(cases),
      new_deaths = deaths - lag(deaths)) %>%
  # Group the data by month 
  group_by(year_month) %>%
  summarize(new_cases = sum(new_cases), new_deaths = sum(new_deaths)) %>%
  ungroup() %>%
  # Plot a line graph with log-scale y-axis.
  ggplot(aes(x = year_month, y = new_cases)) +
  geom_line(aes(color = "new_cases", group = 1)) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths", group = 1)) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = "New COVID19 cases in Canada", y = NULL)
```

The graph displays the monthly new COVID-19 cases starting from February 2020, revealing a pronounced surge during the winter months and a decline during the summer months. The trend stabilized towards the end of 2022 and started to decrease gradually.

## Global trend

To predict a linear model for deaths per thousand and cases per thousand, we calculated these values for each region in the data.

```{r analysis_linear1}
# Create data for every region in global data
global_total <- global %>%
  group_by(Combined_Key) %>%
  # Calculate deaths per thousand and cases per thousand
  summarize(deaths = max(deaths), cases = max(cases),
           Population = max(Population),
           cases_per_thou = 1000 * cases / Population,
           deaths_per_thou = 1000 * deaths / Population) %>%
  filter(cases > 0, Population > 0)

# Create a linear model
mod <- lm(deaths_per_thou ~ cases_per_thou, data = global_total)

# Show summary of the model
summary(mod)

```

Utilize the linear model generated to forecast deaths per thousand based on cases per thousand, subsequently comparing the results with the actual values.
```{r analysis_linear2}
# Predict data using model
global_total_w_pred <- global_total %>% mutate(pred = predict(mod))

# Plot the graph
global_total_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou, color = "actual")) +
  geom_point(aes(x = cases_per_thou, y = pred, color = "predicted")) +
  theme(legend.position = "bottom")
```
The graph illustrates a correlation between deaths per thousand and cases per thousand, exhibiting a near-linear pattern.

## Summary
The report presents an analysis of COVID-19 data for Canada, including trends in cumulative cases and fatalities for the country and its provinces. The analysis includes a calculation of new cases on a monthly basis, demonstrating a surge during the winter months and a decline during the summer months. Additionally, the report examines the global trend and predicts a linear model for deaths per thousand based on cases per thousand, finding a correlation between the two.

# Bias Identification

Using a single data source in the report may limit its ability to capture a comprehensive picture of the COVID-19 situation and result in potential bias. The report also lacks mention of any limitations or sources of error in the data or analysis; and fails to thoroughly examine the geographic location and demographic variables, which may impact the pandemic's patterns and trends. Addressing these shortcomings is necessary for ensuring the report's validity and credibility.


```{r}
sessionInfo()
```
